---
title: "Coin flipping example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{coin-example}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(treepplr)
library(dplyr)
library(ggplot2)
```

## Load model and data files

Load the coin model and example data available within *treepplr.*

```{r}
model <- tp_model("coin")
data <- tp_data("coin")
```

The data in this example is a sequence of coin flip results. *treeppl* can only read data in JSON format, that's why all example data are in this format.

```{r}
str(data)
```

## Run TreePPL

Now we can compile and run the TreePPL program. The function *tp_treeppl()* has many optional arguments to change the inference method used. Here, we will use the default settings and only pass the model and the data.

```{r, eval=FALSE}
output_list <- tp_treeppl(model = model, data = data)
```

```{r, echo=FALSE}
output_list <- readRDS("rdata/coin/output_coin.rds")
```

## Plot the posterior distribution

TreePPL outputs the log weight of each sample, so first we need to get the normalized weights and then we can plot the posterior distribution produced.

```{r, fig.height=5, fig.width=5}
# turn list into a data frame where each row represents one sample 
# and calculate normalized weights from log weights
output <-  tp_parse(output_list) %>% 
  dplyr::mutate(weight = exp(log_weight - max(.$log_weight)))

ggplot(output, aes(samples, weight=weight)) +
  geom_histogram(aes(y = after_stat(density)), 
                 col = "white", fill = "lightblue", binwidth=0.04) +
  geom_density() +
  theme_bw()

```
