---
title: "Coin flipping example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{coin-example}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)

options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup}
library(treepplr)
library(dplyr)
library(ggplot2)
```

## Load model and data files

Load the coin model and example data available within *treepplr.*

```{r}
model <- tp_model("coin")
data <- tp_data("coin")
```

The data in this example is a sequence of coin flip results. *treeppl* can only read data in JSON format, that's why all example data are in this format.

```{r}
str(data)
```

## Run TreePPL

Now we can compile and run the TreePPL program. The function *tp_treeppl()* has many optional arguments to change the inference method used. Here, we will use the default settings for inference methods, run 100 sweeps and 100 particles within each sweep.

```{r, eval=FALSE}
output_list <- tp_treeppl(model = model, data = data, samples = 100, n_runs = 100)
```

```{r, echo=FALSE}
output_list <- readRDS("rdata/coin/output_coin.rds")
```

## Plot the posterior distribution

Just as we compare different MCMC runs/chains to test for convergence, in SMC we
tun several sweeps and then compare their normalizing constants to test if they
converged to the same posterior distribution. The criterion for convergence is 
that the variance of the normalizing constants across all sweeps is lower than 1. 

```{r, fig.height=5, fig.width=5}
tp_smc_convergence(output_list)

output <-  tp_parse(output_list) %>% 
  dplyr::mutate(total_lweight = log_weight + norm_const) %>% 
  dplyr::mutate(norm_weight = exp(total_lweight - max(.$total_lweight)))

ggplot2::ggplot(output, ggplot2::aes(samples, weight = norm_weight)) +
  ggplot2::geom_histogram(ggplot2::aes(y = ggplot2::after_stat(density)),
                 col = "white", fill = "lightblue", binwidth=0.01) +
  ggplot2::geom_density() +
  ggplot2::theme_bw()

```
