---
title: "Constant rate birth (CRB) model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Constant-Rate-Birth}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(treepplr)
library(ape)
library(ggplot2)
library(magrittr)
```


## Data

We will use an example (random) tree that comes with the package.

```{r, fig.width=6, fig.height=4}
tree <- ape::read.tree(system.file(
  "extdata/crb_tree_15_tips.tre", package = "treepplr"))
ape::plot.phylo(tree, cex = 0.5)
```

We need to convert the tree to a **TreePPL** readable format and read the CRB model.

```{r}
data <- tp_phylo_2_TreePPL(tree, age="down-top")
model <- tp_model(system.file("extdata/crb.tppl", package = "treepplr"))
```


## Run treeppl

Compile and run the TreePPL program with standard inference settings.

```{r, eval=FALSE}
output_list <- tp_treeppl(model = model, data = data)
```

```{r, echo=FALSE}
output_list <- readRDS("rdata/crb/output_crb.rds")
```


## Plot posterior

TreePPL outputs the log weight of each sample, so first we need to get the normalized weights and then we can plot the posterior distribution produced.

```{r, fig.height=4, fig.width=6}
# turn list into a data frame where each row represents one sample 
# and calculate normalized weights from log weights
output <-  tp_parse(output_list) %>% 
  dplyr::mutate(weight = exp(log_weight - max(.$log_weight)))

ggplot2::ggplot(output) +
  ggplot2::geom_histogram(aes(samples, y = after_stat(density), weight=weight), 
                          col = "white", fill = "lightblue") +
  ggplot2::geom_density(aes(samples, weight=weight)) +
  ggplot2::theme_bw()

```

