---
title: "Host repertoire model example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{hostrep-example}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE
)
```

```{r, echo=FALSE, message=FALSE}
library(treepplr)
library(ggplot2)
library(magrittr)
library(dplyr)

if(!require("evolnets", quietly = TRUE)) {
 library(devtools)
 if (!require("BiocManager", quietly = TRUE))
   install.packages("BiocManager")
 library(BiocManager)
 BiocManager::install("ggtree")
 BiocManager::install("treeio")
 devtools::install_github("maribraga/evolnets")
 library(ggtree)
 library(treeio)
 library(evolnets)
} else {
 library(evolnets)
}

```

```{r, echo=FALSE}
load("rdata/hostrep/rdata.RData")
```

This is a very simplified workflow in *treepplr* for analysis of host repertoire evolution to show the overall structure. A real analysis would include many other steps such as testing of different inference methods and check of convergence.

## Run treeppl

Run the `hostrep.tppl` TreePPL program using the input data available in the package and default inference options.

```{r, eval=FALSE}
model <- tp_model("hostrep3states")
data <- tp_data("hostrep3states")
output_json <- tp_treeppl(model = model, model_file_name = "hostrep", 
                          data = data, data_file_name = "hostrep")
```

```{r}
output <- tp_parse(output_json)
```

The output from *tp_parse()* contains all information outputted by treeppl, so we need to separate the sampled character histories from the sampled parameter values.

## Parameter estimates

```{r}
logs <-  output[[1]] %>% 
  dplyr::select(iteration, log_weight, mu, beta) %>% 
  unique() %>% 
  dplyr::mutate(weight = exp(log_weight - max(.$log_weight)))

ggplot(logs) +
  geom_histogram(aes(mu, y = after_stat(density), weight=weight), col = "white", fill = "lightblue") +
  geom_density(aes(mu, weight=weight)) +
  theme_bw()

ggplot(logs) +
  geom_histogram(aes(beta, y = after_stat(density), weight=weight), col = "white", fill = "lightblue") +
  geom_density(aes(beta, weight=weight)) +
  theme_bw()
```


## Character history

```{r}
# post-treatment function
get_history <- function(parsed_output){
  
  table <- parsed_output[[1]] %>% 
    dplyr::select(-c(log_weight, mu, beta, lambda1, lambda2, lambda3, lambda4)) %>%
    dplyr::mutate(transition_type = "anagenetic") %>% 
    dplyr::mutate(node_index = dplyr::case_when( # fix to mismatch and 0- to 1-base
                                        node_index == 3 ~ 5, 
                                        node_index == 4 ~ 4,
                                        TRUE ~ node_index + 1)) %>% 
    dplyr::filter(!is.na(log_norm_const))
    
  return(table)
}
```

Extract table with character history samples

```{r}
tp_hist <- get_history(output)
head(tp_hist)
```

This table can be used by *evolnets* for plotting results, together with the phylogenetic trees and the known extant interactions.

```{r, eval=FALSE}
# get data from treepplr
symbiont_tree <- evolnets::read_tree_from_revbayes(
  system.file("extdata/tree_subroot_Rev.tre", package = "treepplr"))
host_tree <- treeio::read.tree(
  system.file("extdata/host_tree.tre", package = "treepplr"))
matrix <- read.csv(
  system.file("extdata/host_matrix.csv", package = "treepplr"), 
  row.names = 1) %>% 
  as.matrix()
```

```{r}
# calculate posterior at nodes
tp_at_nodes <- evolnets::posterior_at_nodes(tp_hist, symbiont_tree, host_tree, state = c(1,2))

# plot data and inferred ancestral states
# fundamental host repertoire
(tp_asr_fund <- evolnets::plot_matrix_phylo(matrix, tp_at_nodes, symbiont_tree, 
                                            host_tree, type = "repertoires", 
                                            repertoire = "fundamental"))
# realized host repertoire
(tp_asr_real <- evolnets::plot_matrix_phylo(matrix, tp_at_nodes, symbiont_tree, 
                                            host_tree, type = "repertoires", 
                                            repertoire = "realized"))
```
