---
title: "Host repertoire model example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{hostrep-example}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

```{r, message=FALSE}
library(treepplr)
library(ggplot2)
library(magrittr)
library(dplyr)
library(ape)

if(!require("evolnets", quietly = TRUE)) {
 library(devtools)
 if (!require("BiocManager", quietly = TRUE))
   install.packages("BiocManager")
 library(BiocManager)
 BiocManager::install("ggtree")
 devtools::install_github("maribraga/evolnets")
 library(ggtree)
 library(evolnets)
} else {
 library(evolnets)
}

```

This is a very simplified workflow in **treepplr** for analysis of host repertoire evolution. The data used here is simulated and the inference is too simple. A real analysis would include many other steps such as testing of different inference methods and check of convergence.

The purpose of this vignette is to show how to run **TreePPL** with **treepplr**, and how to process the output with **evolnets**.

## Load model and data files

Load the 3-state host repertoire model and example data available within **treepplr**.

```{r}
model <- tp_model("hostrep3states")
data <- tp_data("hostrep3states")
```

## Run treeppl

<!-- Now we can compile and run the TreePPL program using the default inference options. -->

<!-- ```{r, eval=FALSE} -->
<!-- output_list <- tp_treeppl(model = model, model_file_name = "hostrep",  -->
<!--                           data = data, data_file_name = "hostrep",  -->
<!--                           samples = 100, n_runs = 5) -->
<!-- ``` -->

<!-- ```{r, echo=FALSE} -->
<!-- output_list <- readRDS("rdata/hostrep/output_hostrep3states.rds") -->
<!-- ``` -->

<!-- ```{r} -->
<!-- output <- tp_parse_host_rep(output_list) -->
<!-- ``` -->

<!-- The output from *tp_parse_host_rep()* contains all information outputted by treeppl, so we need to separate the sampled character histories from the sampled parameter values. -->

<!-- ## Parameter estimates -->

<!-- ```{r} -->
<!-- logs <-  output[[1]] %>%  -->
<!--   dplyr::select(iteration, log_weight, mu, beta) %>%  -->
<!--   unique() %>%  -->
<!--   dplyr::mutate(weight = exp(log_weight - max(.$log_weight))) -->

<!-- ggplot(logs, aes(mu, weight=weight)) + -->
<!--   geom_histogram(aes(y = after_stat(density)), col = "white", fill = "lightblue", binwidth=0.04) + -->
<!--   geom_density() + -->
<!--   theme_bw() -->

<!-- ggplot(logs, aes(mu, weight=weight)) + -->
<!--   geom_histogram(aes(y = after_stat(density) ), col = "white", fill = "lightblue", binwidth=0.04) + -->
<!--   geom_density() + -->
<!--   theme_bw() -->
<!-- ``` -->

<!-- ## Character history -->

<!-- Extract table with character history samples -->

<!-- ```{r} -->
<!-- # post-treatment function -->
<!-- get_history <- function(parsed_output){ -->

<!--   table <- parsed_output[[1]] %>%  -->
<!--     dplyr::select(-c(log_weight, mu, beta, lambda1, lambda2, lambda3, lambda4)) %>% -->
<!--     dplyr::mutate(transition_type = "anagenetic") %>%  -->
<!--     dplyr::mutate(node_index = dplyr::case_when( # fix to mismatch and 0- to 1-base -->
<!--                                         node_index == 3 ~ 5,  -->
<!--                                         node_index == 4 ~ 4, -->
<!--                                         TRUE ~ node_index + 1)) %>%  -->
<!--     dplyr::filter(!is.na(log_norm_const)) -->

<!--   return(table) -->
<!-- } -->

<!-- tp_hist <- get_history(output) -->
<!-- ``` -->


<!-- This table can be used by **evolnets** for plotting results, together with the phylogenetic trees and the known extant interactions. -->

<!-- ```{r} -->
<!-- # get data from treepplr -->
<!-- symbiont_tree <- read_tree_from_revbayes( -->
<!--   system.file("extdata/hostrep3states_treeRev.tre", package = "treepplr")) -->
<!-- host_tree <- read.tree( -->
<!--   system.file("extdata/hostrep3states_host_tree.tre", package = "treepplr")) -->
<!-- matrix <- read.csv( -->
<!--   system.file("extdata/hostrep3states_extant_interactions.csv", package = "treepplr"),  -->
<!--   row.names = 1) %>%  -->
<!--   as.matrix() -->
<!-- ``` -->

<!-- ### Plot data and inferred ancestral fundamental host repertoires -->

<!-- ```{r, fig.height=4, fig.width=6} -->
<!-- # calculate posterior at nodes -->
<!-- tp_at_nodes <- posterior_at_nodes(tp_hist, symbiont_tree, host_tree, state = c(1,2)) -->

<!-- #plot -->
<!-- (tp_asr_fund <- plot_matrix_phylo(matrix, tp_at_nodes, symbiont_tree,  -->
<!--                                             host_tree, type = "repertoires",  -->
<!--                                             repertoire = "fundamental")) -->
<!-- ``` -->

<!-- ### Plot data and inferred ancestral realized host repertoires -->

<!-- ```{r, fig.height=4, fig.width=6} -->
<!-- (tp_asr_real <- plot_matrix_phylo(matrix, tp_at_nodes, symbiont_tree,  -->
<!--                                             host_tree, type = "repertoires",  -->
<!--                                             repertoire = "realized")) -->
<!-- ``` -->

