---
title: 'TreePPL in R: Writing and Running Programs'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tuto}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)

options(rmarkdown.html_vignette.check_title = FALSE)
```

Let’s load the treepplr, TreePPL and all others relevant packages on
this R Markdown document. Markdown is a simple formatting syntax for
authoring HTML, PDF, and MS Word documents. For more details on using R
Markdown see <http://rmarkdown.rstudio.com>.

```{r setup, echo=FALSE}
library(treepplr)
library(dplyr)
library(ggplot2)
```

## Introduction

`treepplr` is an interface for using the TreePPL program. All functions
start with `tp_` to easily distinguish them from functions from other
packages.

For example, the following cell demonstrates a simple TreePPL program
for simulating the flip of a fair coin:

```{r}
unfair_coin_model <- 
  "model function flip() => Bool {
  assume p ~ Bernoulli(0.7);
  return p;
}"
no_data <- tp_data("coin") #ignore this line
exe_path <- tp_compile(model = unfair_coin_model, 
                       method = "smc-apf", 
                       particles = 10)
```

In this example, a treeppl model is created, and the program is
compiled. The argument particles=10 specifies the number of samples to
generate when the program is executed.

To run the TreePPL program, call the function `tp_run`. This executes
the program and returns result in a structured list, which includes a
samples attribute containing the generated samples. While samples may
have different weights in more complex programs, they are equally
weighted in this simple example. We will cover programs with weighted
samples later.

Here’s an example of how to use the compiled program:

```{r, fig.height=5, fig.width=5}
output_list <- tp_run(compiled_model = exe_path, data = no_data, n_sweeps = 1)
result <- data.frame(valeur = unlist(output_list[[1]]$samples))
ggplot(result, aes(x = valeur, fill = valeur)) +
  geom_bar(width = 0.5) +
  scale_fill_manual(values = c("TRUE" = "#2ecc71", "FALSE" = "#e74c3c")) +
  theme_minimal() +
  labs(title = "Coin repartion", x = "Values", y = "Count")
```

## Simple inference

Small setup to make reading easy
```{r}
myplot <- function(df) {
  ggplot(df, aes(x = samples, weight = weights)) +
    geom_histogram(aes(y = after_stat(density)), 
                   bins = 100, fill = "skyblue", color = "white") +
    geom_density(color = "blue", linewidth = 1) +
    theme_minimal()
}

compile_run_format <- function(model, data, particules) {
  exe_path <- tp_compile(model = model, 
                       method = "smc-apf", 
                       particles = particules)
  res <- tp_run(compiled_model = exe_path, data = data, n_sweeps = 1)

  result <- data.frame(samples=unlist(res[[1]]$samples), 
                     weights=exp(as.numeric(unlist(res[[1]]$weights))))
  result
}
```

### Coin model example

```{r}
coin1 <- 
  "model function coin1() => Real {
    assume p ~ Beta(2.0,2.0);
    return p;
}"

result <- compile_run_format(coin1, no_data, 30000)

myplot(result)
```

```{r}
coin2 <- 
  "model function coin2() => Real {
    assume p ~ Beta(2.0,2.0);
    observe true ~ Bernoulli(p);
    return p;
}"

result <- compile_run_format(coin2, no_data, 30000)

myplot(result)
```

```{r}
coin3 <- 
  "model function coin3() => Real {
    assume p ~ Beta(2.0,2.0);
    observe true ~ Bernoulli(p);
    observe true ~ Bernoulli(p);
    observe true ~ Bernoulli(p);
    observe false ~ Bernoulli(p);
    observe true ~ Bernoulli(p);
    observe true ~ Bernoulli(p);
    return p;
}"

result <- compile_run_format(coin3, no_data, 30000)

myplot(result)
```

### Generalized Coin model

```{r}
coin <- 
"model function coin(outcomes: Bool[]) => Real {
  assume p ~ Uniform(0.0, 1.0);
  for i in 1 to (length(outcomes)) {
    observe outcomes[i] ~ Bernoulli(p);
  }
  return p;
}"

data = tp_data(list(outcomes=c(FALSE, TRUE, TRUE, FALSE, TRUE, FALSE, 
                               FALSE, TRUE, TRUE, FALSE, TRUE, FALSE, 
                               TRUE, FALSE, FALSE)))

result <- compile_run_format(coin, data, 50000)

myplot(result)

```

```{r}
foo <- 
"model function foo() => Real {
  assume choice ~ Bernoulli(0.7);
  if(choice){
    assume p ~ Gaussian(-2.0, 1.0);
    return p;
  }
  else{
    assume p ~ Gaussian(3.0, 1.0);
    return p;
  }
}"

result <- compile_run_format(foo, no_data, 40000)

myplot(result)

```

### Task

Try to run the model above using less data, and with another prior, e.g., 
using Beta. How is the change of prior affecting the result? 
What is the effect of the prior if you have a large number of observations?

## Exercise 1: Recursively Defined Models

**Geometric distribution:** X number of Bernoulli trials needed to get one 
success.\
**Parameter:** probability of success (0 < p ≤ 1)\
**Support:** x number of trials, (x ∈ {1,2,3,…})

```{r}
geometric <- 
"model function geometric(p : Real) => Int {
  assume cont ~ Bernoulli(p);
  if cont {
    return 1;
  }
  return 1 + geometric(p);
}"

data = tp_data(list(p=0.5))

result <- compile_run_format(geometric, data, 10000)

ggplot(result, aes(x = samples, weight = weights)) +
    geom_histogram(bins = 100, fill = "skyblue", color = "white") +
    theme_minimal()
```

## Exercise 2: A Graphical Model

$\alpha \sim \text{Exponential}(1)$\
$\beta \sim \text{Gamma}(0.1, 1)$\
$\theta_i \sim \text{Gamma}(\alpha, \beta)\quad i = 1, \ldots, N$\
$y_i \sim \text{Poisson}(\theta_i t_i) \quad i = 1, \ldots, N$

Implement this graphical model and infer $\alpha$.

```{r}
pump <- 
"model function pump(y : Int[], t : Real[]) => Real {
  assume alpha ~ Exponential(1.0);
  assume beta ~ Gamma(0.1, 1.0);
  for i in 1 to (length(y)) {
    assume theta ~ Gamma(alpha, beta);
    observe y[i] ~ Poisson(theta * t[i]);
  }
  return alpha;
}"

data = tp_data(list(y=c(5, 1, 14, 3, 19, 1, 1, 4, 22),
                    t=c(94.3, 15.7, 62.9, 126, 5.24, 31.4, 1.05, 2.1, 10.5)))

result <- compile_run_format(pump, data, 1000000)

myplot(result)
```

## Exercise 3: Bayesian Linear Regression

Suppose you have the following data: `length = [55, 57, 52, 64, 53, 64]` 
and `age = [4, 10, 2, 17, 6, 20]`, where the tuple `(length[i], age[i])` 
represents the length in cm and age inweeks for a baby. 
Create a WebPPL/Stan script that infers a posterior distribution over the 
length of six months old babies by using Bayesian linear regression of the 
form $l_i \sim N(\alpha + \beta a_i, \gamma)$, where $N$ is the 
normal distribution, $l_i$ the length, $a_i$ the age, and $\alpha$, $\beta$ 
and $\gamma$ are random variables.

```{r}
linreg <- 
"type LinRes =
  | LinRes {alpha : Real, beta : Real, gamma : Real}

model function linreg(lengths : Real[], ages : Real[]) => LinRes {
  assume alpha ~ Uniform(0.0, 200.0);
  assume beta ~ Uniform(0.0, 10.0);
  assume gamma ~ Uniform(0.0, 100.0);
  for i in 1 to (length(ages)) {
    observe lengths[i] ~ Gaussian(alpha + beta * ages[i], gamma);
  }
  return LinRes {alpha = alpha, beta = beta, gamma = gamma};
}"

data = tp_data(list(lengths = c(55, 57, 52, 64, 53, 64),
                    ages = c(4, 10, 2, 17, 6, 20)))
particules <- 100000
exe_path <- tp_compile(model = linreg, 
                       method = "smc-apf", 
                       particles = particules)
res <- tp_run(compiled_model = exe_path, data = data, n_sweeps = 1)

extract_info <- function(result, value) {
  unlist(lapply(
  seq(1,length(res[[1]]$samples)), 
  function(x) {
    tmp <- res[[1]]$samples[[x]]["__data__"]$`__data__`
    tmp[value]
  }
 ))
}

df_alpha <- data.frame(samples=extract_info(res, "alpha"),
                 weights=exp(as.numeric(unlist(res[[1]]$weights))))

myplot(df_alpha)

df_beta <- data.frame(samples=extract_info(res, "beta"),
                 weights=exp(as.numeric(unlist(res[[1]]$weights))))

myplot(df_beta)

df_gamma <- data.frame(samples=extract_info(res, "gamma"),
                 weights=exp(as.numeric(unlist(res[[1]]$weights))))

myplot(df_gamma)
```
